Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is an extended abstract of \citet{CompetingBandits-merged}, which in turn subsumes conference papers \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}.}

### 2020-09-04 journal paper/content/sec-intro.tex

The main high-level question is: 
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are the data feedback loops 
%--- when more data leads to more users, which leads to even more data, etc. --- 
and how they relate to the anti-trust considerations?
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.
\asedit{We offer a mix of theoretical results and numerical simulations, in which we study complex interactions between platforms' learning dynamics and users' self-interested behavior.}

@@

%We investigate these questions with
\asedit{
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) compete for users (\emph{agents}). Principals compete on quality rather than on prices, \gaedit{and, crucially, need to engage in exploration in order to learn which actions lead to high quality products}. A new agent arrives and chooses a principal according to some decision rule. The principal selects an action (\eg a list of web search results), the user experiences it and reports a reward. Each principal only observes its own users. Principals commit to their strategies in advance, with \gaedit{the} objective to maximize the market share. We investigate several model variants, %within this framework,
where we vary agents' decision rule or allow a first-mover advantage. %to one of the principals.
%In all variants, agents have little or no information about other agents' choices and rewards.

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a bandit algorithm; as such, it can dynamically adjust itself to the observed rewards, but not to the competition. This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning. Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical. We draw on the machine learning literature to compare bandit algorithms to one another in the absence of competition, so that we can talk about \emph{better} bandit algorithms in a principled way. 
%\footnote{Such comparisons are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Appendix~\ref{app:bg}. Generally,  ``better" algorithms are better in the long run, but could be worse initially. }
One baseline is algorithms that do not explicitly explore (\emph{greedy algorithms}); in the absence of competition, they are known to be terrible for a wide variety of problem instances.}

@@

\xhdr{Interpretation: the inverted-U relationship.}
\asedit{Our findings can be framed in terms of the inverted-U relationship between competition and innovation. This is a well-established concept, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better \citep[\eg][]{aghion2005competition,Vives-08}. We interpret innovation as the adoption of better algorithms,%
\footnote{Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms are well-known and/or similar technologies already exist elsewhere  \citep[\eg see][]{DS-arxiv}.}
%, even if the algorithms themselves are well-known in the research literature }
and control the severity of the competition by varying the agents' decision rule from \HardMax (cut-throat competition) to \HardMaxRandom to \SoftMaxRandom and all the way to the uniform selection.%
\footnote{Agents' decision rule also controls the agents' rationality. While agents' rationality and severity of competition are often modeled separately, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}
We also find another, technically different, inverted-U relationship which zeroes in on the \HardMaxRandom model. 
%We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.
While the inverted-U relationships in prior work focus on the tradeoff between the benefits of the technology and its R\&D costs, in our work the barriers for innovations arise entirely from the reputational consequences of exploration in competition, even in the absence of R\&D costs.}

@@

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of observable data. Approaches such as \citet{lambrecht2015can, bajari2018impact, varian2018artificial} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly modeling the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified by \asedit{the increased \emph{quality} of data gathered by better learning algorithms, not just the quantity thereof.}

@@

\asedit{Our theory takes a basic Bayesian approach, a standard perspective in economic theory, and discovers several strong asymptotic results. Much of the difficulty, both conceptual and technical, is in setting up the model and the theorems. Apart from zeroing in on the \TheoryModel, it was crucial to interpret the results and intuitions from the literature on multi-armed bandits so as to formulate meaningful and productive assumptions on bandit algorithms and Bayesian priors.

@@

One model we studied is suitable for analysis and another for simulations, but not vice versa. A natural implementation of the \TheoryModel requires running time quadratic in the number of rounds,%
\footnote{\Eg this is because at each round $t$, one needs to recompute, and integrate over, a discrete distribution with $t$ possible values, namely the number of agents that have chosen principal $1$ so far.}
which precludes experiments at a sufficient scale. The \ExptsModel features an intricate feedback loop between algorithms' performance, their reputations and agents' choices, which simplifies the simulations but does not appear analytically tractable.}

>>>>

The main high-level question is:
%that we focus on in this paper is:
%\begin{align}\label{eq:main-Q}
\textbf{How does competition incentivize the adoption of better exploration algorithms?}
%\end{align}
This translates into a number of more concrete questions. While it is commonly assumed that better technology always helps, is this so under competition? Does increased competition lead to higher consumer welfare? How significant are the data feedback loops
%--- when more data leads to more users, which leads to even more data, etc. ---
and how they relate to the anti-trust considerations?
%Finding formalizations that admit meaningful answers is a major part of the overall challenge.
We offer a mix of theoretical results and numerical simulations, in which we study complex interactions between platforms' learning dynamics and users' self-interested behavior.

@@

%We investigate these questions with
\xhdr{Our model: competition game.} We consider a stylized duopoly model in which two firms (\emph{principals}) compete for users (\emph{agents}). Principals compete on quality rather than on prices, and engage in exploration in order to learn which actions lead to high quality products. A new agent arrives and chooses a principal according to some decision rule. The principal selects an action (\eg a list of web search results), the user experiences it and reports a reward. Each principal only observes its own users. Principals commit to their strategies in advance, with the objective to maximize the market share. We investigate several model variants, %within this framework,
where we vary agents' decision rule or allow a first-mover advantage. %to one of the principals.
%In all variants, agents have little or no information about other agents' choices and rewards.

Each principal faces a basic and well-studied version of the multi-armed bandit problem, where each reward is drawn independently from a fixed, action-specific distribution. Each principal's strategy is a bandit algorithm; as such, it can dynamically adjust itself to the observed rewards, but not to the competition. This modeling choice reflects the reality of industrial applications, which build on a huge body of knowledge in machine learning. Responding to the competition is another layer of complexity which has not been previously studied in this context, let alone made practical. We draw on the machine learning literature to compare bandit algorithms to one another in the absence of competition, so that we can talk about \emph{better} bandit algorithms in a principled way.
%\footnote{Such comparisons are somewhat subtle, as some algorithms may be better for some problem instances and/or time intervals, and worse for some others. More on this in Appendix~\ref{app:bg}. Generally,  ``better" algorithms are better in the long run, but could be worse initially. }
One baseline is algorithms that do not explicitly explore (\emph{greedy algorithms}); in the absence of competition, they are known to be terrible for a wide variety of problem instances.

\newpage

@@

\xhdr{Interpretation: the inverted-U relationship.}
Our findings can be framed in terms of the inverted-U relationship between competition and innovation. This is a well-established concept, dating back to \cite{Schumpeter-42}, whereby too little or too much competition is bad for innovation, but intermediate levels of competition tend to be better \citep[\eg][]{aghion2005competition,Vives-08}. We interpret innovation as the adoption of better algorithms,%
\footnote{Adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms are well-known and/or similar technologies already exist elsewhere  \citep[\eg see][]{DS-arxiv}.}
%, even if the algorithms themselves are well-known in the research literature
and control the severity of the competition by varying the agents' decision rule from \HardMax (cut-throat competition) to \HardMaxRandom to \SoftMaxRandom and all the way to the uniform selection.%
\footnote{Agents' decision rule also controls the agents' rationality. While agents' rationality and severity of competition are often modeled separately, it is not unusual to have them modeled with the same ``knob" \cite[\eg][]{Gabaix-16}.}
We also find another, technically different, inverted-U relationship which zeroes in on the \HardMaxRandom model.
%We vary rationality/competitiveness inside this model, and track the marginal utility of switching to a better algorithm.
While the inverted-U relationships in prior work focus on the tradeoff between the benefits of the technology and its R\&D costs, in our work the barriers for innovations arise entirely from the reputational consequences of exploration in competition, even in the absence of R\&D costs.

@@

Our results highlight that understanding the performance of learning algorithms in isolation does not necessarily translate to understanding their impact in competition, precisely due to the fact that competition leads to the endogenous generation of observable data. Approaches such as \citet{lambrecht2015can, bajari2018impact, varian2018artificial} argue that the diminishing returns to scale and scope of data in isolation mitigate such data feedback loops,
%as non-existent
but ignore the differences induced by learning in isolation versus under competition. Furthermore, explicitly modeling the interaction between learning technology and data creation allows us to speak on how data advantages are characterized and amplified by the increased \emph{quality} of data gathered by better learning algorithms, not just the quantity thereof.

@@

Our theory takes a basic Bayesian approach, a standard perspective in economic theory, and discovers several strong asymptotic results. Much of the difficulty, both conceptual and technical, is in setting up the model and the theorems. Apart from zeroing in on the \TheoryModel, it was crucial to interpret the results and intuitions from the literature on multi-armed bandits so as to formulate meaningful and productive assumptions on bandit algorithms and Bayesian priors.

@@

One model we studied is suitable for analysis and another for simulations, but not vice versa. A natural implementation of the \TheoryModel requires running time quadratic in the number of rounds,%
\footnote{\Eg this is because at each round $t$, one needs to recompute, and integrate over, a discrete distribution with $t$ possible values, namely the number of agents that have chosen principal $1$ so far.}
which precludes experiments at a sufficient scale. The \ExptsModel features an intricate feedback loop between algorithms' performance, their reputations and agents' choices, which simplifies the simulations but does not appear analytically tractable.
